---
title: Comparing behavioral flexibility across species. Does it work differently in species that rapidly expand their geographic range?
author: 
- '[Logan CJ](http://CorinaLogan.com)^1^'
- '[Deffner D](https://www.eva.mpg.de/ecology/staff/dominik-deffner/index.html)^1^'
date: '`r Sys.Date()`'
output:
  html_document: 
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
    code_folding: hide 
  md_document: 
    toc: true
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
  github_document: 
    toc: true
bibliography: /Users/corina/GitHub/grackles/Files/MyLibrary.bib
csl: /Users/corina/GitHub/grackles/Files/behavioral-ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Make code wrap text so it doesn't go off the page when Knitting to PDF
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

##### Affiliations: 
1) Max Planck Institute for Evolutionary Anthropology

*Corresponding author: corina_logan@eva.mpg.de

***Click [here](https://github.com/corinalogan/flexibility/blob/master/Files/xsppflex.Rmd) for the version-tracked reproducible manuscript (.Rmd file)***

### ABSTRACT

It is generally thought that behavioral flexibility, the ability to change behavior when circumstances change, plays an important role in the ability of a species to rapidly expand their geographic range (e.g., [@lefebvre1997feeding], [@griffin2014innovation], [@chow2016practice], [@sol2000behavioural], [@sol2002behavioural], [@sol2005big], [@sol2007big]). However, it is alternatively possible that an increase in the amount of suitable habitat can be the primary facilitator for a range expansion with behavioral flexibility playing only a minor role. To investigate this possible role of behavioral flexibility across a wide range of taxa, we aim to systematically compare behavioral flexibility in species that do and do not rapidly expand their geographic range. We will collect individual-level data from reversal learning experiments, a standard test of behavioral flexibility, coming from at least XXX different species. Using computational reinforcement-based learning models on those data, we can formally define metrics of behavioral flexibility using 2 key parameters (updating and exploration rate) as well as their temporal dynamics. With those theory-based metrics, we are able to compare flexibility across species without picking up noise that could be due to differences in methodology.Results will elucidate whether rapid geographic range expansions might require behavioral flexibility or whether they are facilitated by the alignment of a species' natural behaviors with an increase in or an introduction to suitable habitat that does not require individuals to alter their behavior to access resources.

### A. STATE OF THE DATA

This preregistration was written (Apr 2020) after beginning to identify parameters involved in behavioral flexibility from two data sets (great-tailed grackles and New Zealand robins), and prior to incorporating any other data sets into the exploration. *[This preregistration was submitted in ??? 2020 to PCI Ecology for pre-study peer review. - Dominik - let's discuss whether we need pre-study peer review on this]*

### B. DEFINING AND MEASURING FLEXIBILITY THROUGH COMPUTATIONAL LEARNING MODELS

Bond and colleagues (2007) identified at least three different, though interrelated, connotations of flexibility in the animal literature. 

1) the ability to rapidly modify behavior based on brief experience,

2) being versatile, exploratory, and playful, modifying behavior even when circumstances do not require it and 

3) the flexibility to change and reverse behavior patterns. 

We aim to formulate a computational learning model that captures all 3 dimensions. Looking to infer cognition (i.e. behavioral flexibility) from behavior, we are faced with a so called inverse problem, i.e. going from (overt) observations to (hidden) causes. This constitutes a problem because typically many different processes can result in the same empirical pattern. By formulating scientific learning models as statistical models, however, we can estimate which parameter values are most compatible with the observed choices. Here, we use multilevel Bayesian reinforcement learning models to analyze data from serial reversal learning experiments. From trial to trial, such models update the latent values of different options based on previous values and recently experienced payoffs. We parameterize our models using two parameters, phi and lambda. Phi describes the weight of recent experience and can be interpreted as the learning or updating rate of an individual. This loosely corresponds to the first and third connotation of Bond et al. (2007), the ability to rapidly and adaptively change behavior in light of new experiences. Lambda represents the exploration rate of an individual and is sometimes also called inverse temperature. This closely corresponds to the second connotation of internally generated behavioral variation, exploration or creativity. Technically, this parameter controls how sensitive choices are to differences in tha latent values of both options. As lambda gets larger, choices become more deterministic, as it gets smaller, choices become more exploratory (random choice if lambda equals 0). 
On top of individual-level parameters, we also aim to include time-varying learning parameters to estimate strategic changes in learning patterns. In particular, flexible animals should be especially exploratory and sensitive to new experiences when reward locations change. After learning the new location, they should explore less and stick with the option that provded successful. Such adaptive temporal dynamics in learning strategies have been observed in human learning experiments, where both phi and lambda changed in the predicted directions after changes in payoffs (Deffner, Kleinow & McElreath, under review).


### C. HYPOTHESES

#### H1: Behavioral flexibility (measured by reversal learning experiments) can be represented as a combination of parameters, such as learning speed and amount of exploration. These parameters can be used to generate a model that compares flexibility across species without picking up noise that could be due to differences in methodology. 

**Prediction 1:** We are able to identify flexibility parameters and test for their reliability using a few data sets from diverse taxa.

**Prediction 1 alternative 1:** We are not able to identify flexibility parameters or test for their reliability using a few data sets from diverse taxa.

**Table 1.** The parameters we explored and what they mean.  

```{r table1, eval=TRUE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}
d <- read.csv(url("https://raw.githubusercontent.com/corinalogan/flexibility/master/Files/xsppflex_Table1.csv"), header=T, sep=",", stringsAsFactors=F) 
dat<-data.frame(d)

library(reactable)
reactable(dat, highlight=TRUE, bordered=FALSE, compact=TRUE, wrap=TRUE, resizable=TRUE)
```

**Table 2.** Differences in reversal learning experimental designs and how these differences affect the robustness of our model.  

```{r table2, eval=TRUE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}
d <- read.csv(url("https://raw.githubusercontent.com/corinalogan/flexibility/master/Files/xsppflex_Table2.csv"), header=T, sep=",", stringsAsFactors=F) 
dat<-data.frame(d)

library(reactable)
reactable(dat, highlight=TRUE, bordered=FALSE, compact=TRUE, wrap=TRUE, resizable=TRUE)
```


#### H2: Those species that have rapidly expanded their geographic range will show differences in their flexibility parameters compared with species that have not rapidly expanded their range.

**Prediction 2:** If a change in behavior is facilitating range expansions, then we predict that species that have undergone a range expansion will learn faster and explore more than species that have not undergone a rapid range expansion. 

**Prediction 2 alternative 1:** If a change in behavior is not the main facilitator of geographic range expansions, then we predict that there will be no differences in the flexibility parameters across species that have undergone a rapid range expansion and those that have not. It is also possible that geographically restricted species may require flexibility to be able to remain in their suitable habitat, thus contributing to the lack of a difference in flexibility parameters.

**Prediction 2 alternative 2:** If a change in behavior is not the main facilitator of geographic range expansions, and if geographically restricted species require flexibility to be able to remain in their suitable habitat, then we predict that species that have not undergone a rapid range expansion will learn faster and explore more than species that have rapidly expanded. *[I need to make this more distinct from P2 alt 1 - or maybe I can't tell them apart...]*


### D. METHODS

#### Planned Sample
	
We will search through the literature for reversal learning experiments and attempt to obtain trial-level data for the individuals in these studies for use in our investigation. If these data are not published, we will contact the authors to request the contribution of their data to this investigation. Our aims are to 1) include data from a diverse range of taxa, including humans, and 2) have a sample where roughly half of the species have rapidly expanded their geographic range while the other half has not.

#### Sample size rationale
	
*[Dominik - can we do some simulations to determine the minimum number of species needed to obtain robust results?]*

#### Data collection stopping rule
	
We will stop collecting data once we have data for 40 species. If we are not able to obtain data for 40 species (because the data are not published, are no longer available, or authors are unwilling to share), we will stop collecting data once we have obtained the data that is the most likely to be obtainable. 

#### Protocols

 - **Identifying species that have undergone a geographic range expansion**: we will use the IUCN Red List of Threatened Species classification for each species to determine whether it has expanded its range (classified as: least concern) or not (i.e., near threatened, vulnerable, endangered, critically endangered, extinct in the wild). *[CL needs to find a database that lists species according to whether they have expanded their range or not. IUCN Red List classifications won't work for this]*
 - **Parameters we investigated**
    - Learning speed:
    - Exploration: 
    - Learning strategy (within individual learning rate changes over time)

#### Open data

The data involved in this study are either already published openly, or obtained directly from the authors. In the latter case, we will attempt to publish the data we use in the Knowledge Network for Biocomplexity's data repository. However, this will depend on the wishes of the authors who own the data.

#### Randomization and counterbalancing 

No randomization or counterbalancing is involved in this investigation.

#### Blinding of conditions during analysis

No blinding is involved in this investigation.

### E. ANALYSIS PLAN

Analyses will be conducted in R (current version `r getRversion()`; @rcoreteam) and Stan (version 2.18, @carpenter2017stan). 

#### H1: identifying parameters

**Table 2.** The species included in our analyses and their associated flexibility parameters (from reversal learning experiments).  

```{r table3, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}
d <- read.csv ("/Users/corina/GitHub/flexibility/Files/xsppflex_Table2.csv", header=F, sep=",", stringsAsFactors=F) 

library(reactable)
reactable(d, highlight=TRUE, bordered=FALSE, compact=TRUE, wrap=TRUE, resizable=TRUE,
          columns = list(
            V1 = colDef(name="Site"),
            V2 = colDef(name="Range position"),
            V3 = colDef(name="Breeding since"),
            V4 = colDef(name="Number of years breeding"),
            V5 = colDef(name="Average number of generations"),
            V6 = colDef(name="Citation")
          ))
```

#### H2: cross-species comparison

*Response variables*

1) Learning rate

2) Exploration rate

3) Temporal dynamics in both

All learning parameters are estimated from a series of computational learning model of increasing complexity. A baseline model will only contain individual-level learning parameters. In a second model, we aim to estimate separate parameters for trials before and after reversal. Lastly, we aim to estimate full temporal trends through Gaussian processes. 

*Explanatory variable*

1) Rapidly expanded its geographic range (yes, no)


### F. ETHICS

This research did not require any permits because it uses data that were already collected as part of other investigations.

### G. AUTHOR CONTRIBUTIONS

**Logan:** Hypothesis development, data collection (from the literature and authors), data analysis and interpretation, write up, revising/editing, materials/funding.

**Deffner:** Hypothesis development, data collection (from the literature and authors), data analysis and interpretation, write up, revising/editing.

### H. FUNDING

This research is funded by the Department of Human Behavior, Ecology and Culture at the Max Planck Institute for Evolutionary Anthropology.

### I. CONFLICT OF INTEREST DISCLOSURE

We, the authors, declare that we have no financial conflicts of interest with the content of this article. CJ Logan is a Recommenders and on the Managing Board at PCI Ecology.

### J. ACKNOWLEDGEMENTS



### J. [REFERENCES](MyLibrary.bib)
